---
title: "Array correlations"
author: "Philipp Ross"
date: 2018-09-25
output: workflowr::wflow_html
editor_options: 
  chunk_output_type: console
---



**Last updated:** `r Sys.Date()`

**Code version:** `r workflowr::extract_commit(".", 1)$sha1`

One of the first things we wanted to check was whether the time points from our RNA-seq data correlated well with published microarray data. First we need to import normalized RNA-seq and microarray data:

```{r}
seq3d7 <- readr::read_tsv("../data/array_correlations/selected_globnorm_3d7_rpkms.txt")
seqhb3 <- readr::read_tsv("../data/array_correlations/selected_globnorm_hb3_rpkms.txt") 
seqit <- readr::read_tsv("../data/array_correlations/selected_globnorm_it_rpkms.txt") 
array3d7 <- readr::read_tsv("../data/array_correlations/selected_3d7array_foldchanges_v3_geneids.txt")
```

Now we can calculate the correlations between all time points for each strain compared to the original 3D7 array data.

## 3D7

```{r}
tmp <- dplyr::inner_join(array3d7, seq3d7, by = c("GeneID"="gene_id"))
corr <- round(cor(na.omit(tmp[2:61])), 2)
ggcorrplot::ggcorrplot(corr, hc.order = FALSE, outline.col = "white")
```

## HB3

```{r}
tmp <- dplyr::inner_join(array3d7, seqhb3, by = c("GeneID"="gene_id"))
corr <- round(cor(na.omit(tmp[2:61])), 2)
ggcorrplot::ggcorrplot(corr, hc.order = FALSE, outline.col = "white")
```

## IT

```{r}
tmp <- dplyr::inner_join(array3d7, seqit, by = c("GeneID"="gene_id"))
corr <- round(cor(na.omit(tmp[2:61])), 2)
ggcorrplot::ggcorrplot(corr, hc.order = FALSE, outline.col = "white")
```

## Session Information

