---
title: "Comparing predicted UTRs"
author: "Philipp Ross"
date: 2018-09-25
output: workflowr::wflow_html
editor_options: 
  chunk_output_type: console
---

```{r,echo=F}
# ---- knitr-opts-chunk ----
knitr::opts_chunk$set(
  comment = NA,
  fig.align = "center",
  tidy = FALSE,
  fig.path = paste0("figure/", knitr::current_input(), "/"),
  warning = FALSE,
  message = FALSE
)

# ---- load-libraries ----
source("../code/utils.R")
load_essentials()
```

```{r}
scatter_theme <- theme_classic() + 
  theme(axis.text.x=element_text(size=16), 
        axis.text.y=element_text(size=16),
        axis.title=element_blank(),
        axis.line.x=element_line(colour="black",size=1.5),
        axis.ticks.x=element_line(colour="black",size=1.5),
        axis.line.y=element_line(colour="black",size=1.5),
        axis.ticks.y=element_line(colour="black",size=1.5),
        legend.text=element_blank())

hist_theme <- theme_classic() + 
  theme(axis.text.x=element_text(size=16), 
        axis.text.y=element_text(size=16),
        axis.title=element_blank(),
        axis.line.x=element_line(colour="black",size=1.5),
        axis.ticks.x=element_line(colour="black",size=1.5),
        axis.line.y=element_line(colour="black",size=1.5),
        axis.ticks.y=element_line(colour="black",size=1.5),
        legend.text=element_blank())

height <- 4
width  <- 4
```

## Comparing different UTR and TSS predictions

Since we were able to predict both UTRs and TSSs using our data, we wanted to know how our predictions compared to previously published predictions. Here, we compare the UTRs predicted in [Caro et al.](https://www.ncbi.nlm.nih.gov/pubmed/25493618) and [Adjalley et al.](https://www.ncbi.nlm.nih.gov/pubmed/26947071).

What we see is that although there are some large deviations, for the majority of 5UTR and TSS predictions, the results are not very different with a mean hovering around zero base pairs of difference between the start positions of our predicted 5' UTRs and TSS and those that were previously published.

First, let's import our 5UTR data:

```{r}
library(org.Pf.plasmo.db)
aliases <- tibble::as_tibble(data.frame(org.Pf.plasmoALIAS2ORF))

fiveutrs <- tibble::as_tibble(rtracklayer::import.gff("../output/final_utrs/final_5utrs_3d7.gff"))
fiveutrs$Parent <- unlist(fiveutrs$Parent)

threeutrs <- tibble::as_tibble(rtracklayer::import.gff("../output/final_utrs/final_3utrs_3d7.gff"))
threeutrs$Parent <- unlist(threeutrs$Parent)
```

## Comparing to Caro et al.

```{r}
# Comparing our UTR estimates to the Derisi predictions

derisi1 <- tibble::as_tibble(rtracklayer::import.bed("../data/compare_utrs/GSM1410291_UTRs_1.bed"))
derisi2 <- tibble::as_tibble(rtracklayer::import.bed("../data/compare_utrs/GSM1410292_UTRs_2.bed"))
derisi3 <- tibble::as_tibble(rtracklayer::import.bed("../data/compare_utrs/GSM1410293_UTRs_3.bed"))
derisi4 <- tibble::as_tibble(rtracklayer::import.bed("../data/compare_utrs/GSM1410294_UTRs_4.bed"))
derisi5 <- tibble::as_tibble(rtracklayer::import.bed("../data/compare_utrs/GSM1410295_UTRs_5.bed"))

fix_derisi_utrs <- function(set) {
  set$name <- stringi::stri_replace_last(set$name,replacement=" ",regex="_")
  set <- set %>% tidyr::separate(name,into = c("gene_id","type"), sep =" ")
  set$gene_id <- toupper(set$gene_id)
  out <- dplyr::inner_join(set, aliases, by=c("gene_id"="alias_symbol"))
  return(out)
}

derisi1 <- fix_derisi_utrs(derisi1)
derisi2 <- fix_derisi_utrs(derisi2)
derisi3 <- fix_derisi_utrs(derisi3)
derisi4 <- fix_derisi_utrs(derisi4)
derisi5 <- fix_derisi_utrs(derisi5)

longest_derisi <- dplyr::bind_rows(derisi1, derisi2, derisi3, derisi4, derisi5) %>% 
  dplyr::group_by(gene_id.y, type) %>% 
  summarize(width=max(width)) %>% 
  dplyr::ungroup()
```

### Longest 5' UTRs

```{r}
tmp1 <- dplyr::filter(longest_derisi, type=="5p") %>% 
  dplyr::select(gene_id.y, width) %>%
  dplyr::rename(derisi_length=width)
tmp2 <- dplyr::select(fiveutrs, Parent, width)
compare_derisi <- dplyr::inner_join(tmp1,tmp2,by=c("gene_id.y"="Parent"))
rm(tmp1,tmp2)

g <- ggplot(compare_derisi,aes(derisi_length+1,width+1)) + 
  geom_point(alpha=0.3,size=3.0) +
  geom_abline(color="red") +
  xlab("Caro Predictions") + 
  ylab("Chappell Predictions") +
  scale_x_continuous(trans="log10",breaks=c(1,10,100,1000,10000),labels=scales::trans_format('log10',math_format(10^.x)),limits=c(1,10000)) +
  scale_y_continuous(trans="log10",breaks=c(1,10,100,1000,10000),labels=scales::trans_format('log10',math_format(10^.x)),limits=c(1,10000)) +
  scatter_theme

ggsave(plot=g,filename="../output/compare/derisi_5utr_scatter.png",height=height,width=width)
ggsave(plot=g,filename="../output/compare/derisi_5utr_scatter.svg",height=height,width=width)
print(g)

g <- ggplot(compare_derisi,aes(derisi_length-width)) +
  geom_histogram(color="grey70",fill="black",bins=25) +
  geom_vline(xintercept=0,color="red",linetype=2,size=1.0) +
  xlab("Frequency") +
  ylab("UTR Length Difference") +
  hist_theme

ggsave(plot=g,filename="../output/compare/derisi_5utr_histogram.png",height=height,width=width)
ggsave(plot=g,filename="../output/compare/derisi_5utr_histogram.svg",height=height,width=width)
print(g)

cor.test(compare_derisi$derisi_length,compare_derisi$width)
compare_derisi %>% dplyr::mutate(diff=derisi_length-width) %>% summary()
```

### Longest 3' UTRs

```{r}
tmp1 <- dplyr::filter(longest_derisi, type=="3p") %>% 
  dplyr::select(gene_id.y, width) %>%
  dplyr::rename(derisi_length=width)
tmp2 <- dplyr::select(threeutrs, Parent, width)
compare_derisi <- dplyr::inner_join(tmp1,tmp2,by=c("gene_id.y"="Parent"))
rm(tmp1,tmp2)

g <- ggplot(compare_derisi,aes(derisi_length+1,width+1)) +
  geom_point(alpha=0.3,size=3.0) +
  geom_abline(color="red") +
  xlab("") + 
  ylab("") +
  scale_x_continuous(trans="log10",breaks=c(1,10,100,1000,10000),
                     labels=scales::trans_format('log10',math_format(10^.x)),limits=c(1,10000)) +
  scale_y_continuous(trans="log10",breaks=c(1,10,100,1000,10000),
                     labels=scales::trans_format('log10',math_format(10^.x)),limits=c(1,10000)) +
  scatter_theme

ggsave(plot=g,filename="../output/compare/derisi_3utr_scatter.png",height=height,width=width)
ggsave(plot=g,filename="../output/compare/derisi_3utr_scatter.svg",height=height,width=width)
print(g)

g <- ggplot(compare_derisi,aes(derisi_length-width)) +
  geom_histogram(color="grey70",fill="black",bins=25) +
  geom_vline(xintercept=0,color="red",linetype=2,size=1.0) +
  xlab("") +
  ylab("") +
  hist_theme

ggsave(plot=g,filename="../output/compare/derisi_3utr_histogram.png",height=height,width=width)
ggsave(plot=g,filename="../output/compare/derisi_3utr_histogram.svg",height=height,width=width)
print(g)

cor.test(compare_derisi$derisi_length,compare_derisi$width)
compare_derisi %>% dplyr::mutate(diff=derisi_length-width) %>% summary()
```

### Timepoint 1

```{r}
tmp1 <- dplyr::filter(derisi1, type=="5p") %>% 
  dplyr::select(gene_id.y, width) %>%
  dplyr::rename(derisi_length=width)
tmp2 <- dplyr::select(utrs, Parent, width)
compare_derisi <- dplyr::inner_join(tmp1,tmp2,by=c("gene_id.y"="Parent"))
rm(tmp1,tmp2)

g <- ggplot(compare_derisi,aes(derisi_length,width)) + 
  geom_point(alpha=0.2) + 
  geom_abline(color="red") +
  xlab("Derisi Predictions") + 
  ylab("Chappell Predictions") 

cowplot::save_plot(filename="../output/compare/derisi_scatter_1.png",plot=g)
cowplot::save_plot(filename="../output/compare/derisi_scatter_1.svg",plot=g)
print(g)

g <- ggplot(compare_derisi,aes(derisi_length-width)) + 
  geom_histogram(color="grey70") +
  xlab("") + 
  ylab("")

cowplot::save_plot(filename="../output/compare/derisi_histogram_1.png",plot=g)
cowplot::save_plot(filename="../output/compare/derisi_histogram_1.svg",plot=g)
print(g)

cor.test(compare_derisi$derisi_length,compare_derisi$width)
compare_derisi %>% dplyr::mutate(diff=derisi_length-width) %>% summary()
```

### Timepoint 2

```{r}
tmp1 <- dplyr::select(derisi2, gene_id.y, width)
tmp2 <- dplyr::select(utrs, Parent, width)
compare_derisi <- dplyr::inner_join(tmp1,tmp2,by=c("gene_id.y"="Parent"))
rm(tmp1,tmp2)

g <- ggplot(compare_derisi,aes(width.x,width.y)) + 
  geom_point() + 
  geom_abline(color="red") +
  xlab("") + 
  ylab("") 

cowplot::save_plot(filename="../output/compare/derisi_scatter_2.png",plot=g)
cowplot::save_plot(filename="../output/compare/derisi_scatter_2.svg",plot=g)
print(g)

g <- ggplot(compare_derisi,aes(width.x-width.y)) + 
  geom_histogram(color="grey70") +
  xlab("") + 
  ylab("")

cowplot::save_plot(filename="../output/compare/derisi_histogram_2.png",plot=g)
cowplot::save_plot(filename="../output/compare/derisi_histogram_2.svg",plot=g)
print(g)

cor.test(compare_derisi$width.x,compare_derisi$width.y)
compare_derisi %>% dplyr::mutate(diff=width.x-width.y) %>% summary()
```

### Timepoint 3

```{r}
tmp1 <- dplyr::select(derisi3, gene_id.y, width)
tmp2 <- dplyr::select(utrs, Parent, width)
compare_derisi <- dplyr::inner_join(tmp1,tmp2,by=c("gene_id.y"="Parent"))
rm(tmp1,tmp2)

g <- ggplot(compare_derisi,aes(width.x,width.y)) + 
  geom_point() + 
  geom_abline(color="red") +
  xlab("") + 
  ylab("") 

cowplot::save_plot(filename="../output/compare/derisi_scatter_3.png",plot=g)
cowplot::save_plot(filename="../output/compare/derisi_scatter_3.svg",plot=g)
print(g)

g <- ggplot(compare_derisi,aes(width.x-width.y)) + 
  geom_histogram(color="grey70") +
  xlab("") + 
  ylab("")

cowplot::save_plot(filename="../output/compare/derisi_histogram_3.png",plot=g)
cowplot::save_plot(filename="../output/compare/derisi_histogram_3.svg",plot=g)
print(g)

cor.test(compare_derisi$width.x,compare_derisi$width.y)
compare_derisi %>% dplyr::mutate(diff=width.x-width.y) %>% summary()
```

### Timepoint 4

```{r}
tmp1 <- dplyr::select(derisi4, gene_id.y, width)
tmp2 <- dplyr::select(utrs, Parent, width)
compare_derisi <- dplyr::inner_join(tmp1,tmp2,by=c("gene_id.y"="Parent"))
rm(tmp1,tmp2)

g <- ggplot(compare_derisi,aes(width.x,width.y)) + 
  geom_point() + 
  geom_abline(color="red") +
  xlab("") + 
  ylab("") 

cowplot::save_plot(filename="../output/compare/derisi_scatter_4.png",plot=g)
cowplot::save_plot(filename="../output/compare/derisi_scatter_4.svg",plot=g)
print(g)

g <- ggplot(compare_derisi,aes(width.x-width.y)) + 
  geom_histogram(color="grey70") +
  xlab("") + 
  ylab("")

cowplot::save_plot(filename="../output/compare/derisi_histogram_4.png",plot=g)
cowplot::save_plot(filename="../output/compare/derisi_histogram_4.svg",plot=g)
print(g)

cor.test(compare_derisi$width.x,compare_derisi$width.y)
compare_derisi %>% dplyr::mutate(diff=width.x-width.y) %>% summary()
```

### Timepoint 5

```{r}
tmp1 <- dplyr::select(derisi5, gene_id.y, width)
tmp2 <- dplyr::select(utrs, Parent, width)
compare_derisi <- dplyr::inner_join(tmp1,tmp2,by=c("gene_id.y"="Parent"))
rm(tmp1,tmp2)

g <- ggplot(compare_derisi,aes(width.x,width.y)) + 
  geom_point() + 
  geom_abline(color="red") +
  xlab("") + 
  ylab("") 

cowplot::save_plot(filename="../output/compare/derisi_scatter_5.png",plot=g)
cowplot::save_plot(filename="../output/compare/derisi_scatter_5.svg",plot=g)
print(g)

g <- ggplot(compare_derisi,aes(width.x-width.y)) + 
  geom_histogram(color="grey70") +
  xlab("") + 
  ylab("")

cowplot::save_plot(filename="../output/compare/derisi_histogram_5.png",plot=g)
cowplot::save_plot(filename="../output/compare/derisi_histogram_5.svg",plot=g)
print(g)

cor.test(compare_derisi$width.x,compare_derisi$width.y)
compare_derisi %>% dplyr::mutate(diff=width.x-width.y) %>% summary()
```

## Comparing to Adjalley et al.

```{r}
# Comparing our UTR estimates to the Adjalley predictions
adjalley <- tibble::as_tibble(rtracklayer::import.gff("../data/compare_utrs/sorted_Adjalley_Chabbert_TSSs.gff")) %>%
  dplyr::filter(FilterSize > 4) %>%
  dplyr::inner_join(dplyr::select(fiveutrs,Parent,end), by=c("AssignedFeat"="Parent")) %>%
  dplyr::select(AssignedFeat,start,end.x,end.y,strand) %>%
  dplyr::rename(end=end.y,end_position=end.x)

# calculate longest predictions

plus <- adjalley %>% 
  dplyr::filter(strand=="+") %>%
  dplyr::mutate(position=(start+end_position)/2,width=end-start) %>%
  dplyr::filter(width>=0) %>%
  dplyr::group_by(AssignedFeat) %>% 
  dplyr::summarise(start=min(position))
minus <- adjalley %>% 
  dplyr::filter(strand=="-") %>% 
  dplyr::mutate(position=(start+end_position)/2,width=start-end) %>%
  dplyr::filter(width>=0) %>%
  dplyr::group_by(AssignedFeat) %>% 
  dplyr::summarise(start=max(position))

longest_adjalley <- dplyr::bind_rows(plus,minus)

# calculate shortest predictions

plus <- adjalley %>% 
  dplyr::filter(strand=="+") %>%
  dplyr::mutate(position=(start+end_position)/2,width=end-start) %>%
  dplyr::filter(width>=0) %>%
  dplyr::group_by(AssignedFeat) %>% 
  dplyr::summarise(start=max(position))
minus <- adjalley %>% 
  dplyr::filter(strand=="-") %>% 
  dplyr::mutate(position=(start+end_position)/2,width=start-end) %>%
  dplyr::filter(width>=0) %>%
  dplyr::group_by(AssignedFeat) %>% 
  dplyr::summarise(start=min(position))

shortest_adjalley <- dplyr::bind_rows(plus,minus)

# calculate mean predictions

plus <- adjalley %>% 
  dplyr::filter(strand=="+") %>%
  dplyr::mutate(position=(start+end_position)/2,width=end-start) %>%
  dplyr::filter(width>=0) %>%
  dplyr::group_by(AssignedFeat) %>% 
  dplyr::summarise(start=mean(position))
minus <- adjalley %>% 
  dplyr::filter(strand=="-") %>% 
  dplyr::mutate(position=(start+end_position)/2,width=start-end) %>%
  dplyr::filter(width>=0) %>%
  dplyr::group_by(AssignedFeat) %>% 
  dplyr::summarise(start=mean(position))

mean_adjalley <- dplyr::bind_rows(plus,minus)

# calculate median predictions

plus <- adjalley %>% 
  dplyr::filter(strand=="+") %>%
  dplyr::mutate(position=(start+end_position)/2,width=end-start) %>%
  dplyr::filter(width>=0) %>%
  dplyr::group_by(AssignedFeat) %>% 
  dplyr::summarise(start=median(position))
minus <- adjalley %>% 
  dplyr::filter(strand=="-") %>% 
  dplyr::mutate(position=(start+end_position)/2,width=start-end) %>%
  dplyr::filter(width>=0) %>%
  dplyr::group_by(AssignedFeat) %>% 
  dplyr::summarise(start=median(position))

median_adjalley <- dplyr::bind_rows(plus,minus)
```

### Longest 5' UTRs

```{r}
tmp1 <- dplyr::select(longest_adjalley, AssignedFeat, start)
tmp2 <- dplyr::select(fiveutrs, Parent, start, end, width, strand)
compare_adjalley <- dplyr::inner_join(tmp1,tmp2,by=c("AssignedFeat"="Parent")) %>%
  dplyr::mutate(adjalley_width=ifelse(strand=="+",end-start.x,start.x-end))

g <- ggplot(compare_adjalley,aes(adjalley_width+1,width+1)) +
  geom_point(alpha=0.3,size=3.0) +
  geom_abline(color="red") +
  xlab("Adjalley Predictions") + 
  ylab("Chappell Predictions") +
  scale_x_continuous(trans="log10",breaks=c(1,10,100,1000,10000),
                     labels=scales::trans_format('log10',math_format(10^.x)),limits=c(1,10000)) +
  scale_y_continuous(trans="log10",breaks=c(1,10,100,1000,10000),
                     labels=scales::trans_format('log10',math_format(10^.x)),limits=c(1,10000)) +
  scatter_theme

cowplot::save_plot(filename="../output/compare/adjalley_longest_scatter.png",plot=g)
cowplot::save_plot(filename="../output/compare/adjalley_longest_scatter.svg",plot=g)
print(g)

g <- ggplot(compare_adjalley,aes(adjalley_width-width)) + 
  geom_histogram(color="grey70",fill="black",bins=25) +
  geom_vline(xintercept=0,color="red",linetype=2,size=1.0) +
  xlab("UTR Length Difference") +
  ylab("Frequency") +
  hist_theme

cowplot::save_plot(filename=paste0("../output/compare/adjalley_longest_histogram.png"),plot=g)
cowplot::save_plot(filename=paste0("../output/compare/adjalley_longest_histogram.svg"),plot=g)
print(g)

cor.test(compare_adjalley$adjalley_width,compare_adjalley$width)
compare_adjalley %>% dplyr::mutate(diff=adjalley_width-width) %>% summary() %>% print()
```

### Shortest 5' UTRs

```{r}
tmp1 <- dplyr::select(shortest_adjalley, AssignedFeat, start)
tmp2 <- dplyr::select(fiveutrs, Parent, start, end, width, strand)
compare_adjalley <- dplyr::inner_join(tmp1,tmp2,by=c("AssignedFeat"="Parent")) %>%
  dplyr::mutate(adjalley_width=ifelse(strand=="+",end-start.x,start.x-end))

g <- ggplot(compare_adjalley,aes(adjalley_width+1,width+1)) +
  geom_point(alpha=0.3,size=3.0) +
  geom_abline(color="red") +
  xlab("") + 
  ylab("") +
  scale_x_continuous(trans="log10",breaks=c(1,10,100,1000,10000,10000),
                     labels=scales::trans_format('log10',math_format(10^.x)),limits=c(1,10000)) +
  scale_y_continuous(trans="log10",breaks=c(1,10,100,1000,10000,10000),
                     labels=scales::trans_format('log10',math_format(10^.x)),limits=c(1,10000)) +
  scatter_theme

cowplot::save_plot(filename="../output/compare/adjalley_shortest_scatter.png",plot=g)
cowplot::save_plot(filename="../output/compare/adjalley_shortest_scatter.svg",plot=g)
print(g)

g <- ggplot(compare_adjalley,aes(adjalley_width-width)) + 
  geom_histogram(color="grey70",fill="black",bins=25) +
  geom_vline(xintercept=0,color="red",linetype=2,size=1.0) +
  xlab("") +
  ylab("") +
  hist_theme

cowplot::save_plot(filename=paste0("../output/compare/adjalley_shortest_histogram.png"),plot=g)
cowplot::save_plot(filename=paste0("../output/compare/adjalley_shortest_histogram.svg"),plot=g)
print(g)

cor.test(compare_adjalley$adjalley_width,compare_adjalley$width)
compare_adjalley %>% dplyr::mutate(diff=adjalley_width-width) %>% summary() %>% print()
```

### Mean 5' UTRs

```{r}
tmp1 <- dplyr::select(mean_adjalley, AssignedFeat, start)
tmp2 <- dplyr::select(fiveutrs, Parent, start, end, width, strand)
compare_adjalley <- dplyr::inner_join(tmp1,tmp2,by=c("AssignedFeat"="Parent")) %>%
  dplyr::mutate(adjalley_width=ifelse(strand=="+",end-start.x,start.x-end))

g <- ggplot(compare_adjalley,aes(adjalley_width+1,width+1)) +
  geom_point(alpha=0.3,size=3.0) +
  geom_abline(color="red") +
  xlab("") + 
  ylab("") +
  scale_x_continuous(trans="log10",breaks=c(1,10,100,1000,10000),
                     labels=scales::trans_format('log10',math_format(10^.x)),limits=c(1,10000)) +
  scale_y_continuous(trans="log10",breaks=c(1,10,100,1000,10000),
                     labels=scales::trans_format('log10',math_format(10^.x)),limits=c(1,10000)) +
  scatter_theme

cowplot::save_plot(filename="../output/compare/adjalley_mean_scatter.png",plot=g)
cowplot::save_plot(filename="../output/compare/adjalley_mean_scatter.svg",plot=g)
print(g)

g <- ggplot(compare_adjalley,aes(adjalley_width-width)) + 
  geom_histogram(color="grey70",fill="black",bins=25) +
  geom_vline(xintercept=0,color="red",linetype=2,size=1.0) +
  xlab("") +
  ylab("") +
  hist_theme

cowplot::save_plot(filename=paste0("../output/compare/adjalley_mean_histogram.png"),plot=g)
cowplot::save_plot(filename=paste0("../output/compare/adjalley_mean_histogram.svg"),plot=g)
print(g)

cor.test(compare_adjalley$adjalley_width,compare_adjalley$width)
compare_adjalley %>% dplyr::mutate(diff=adjalley_width-width) %>% summary() %>% print()
```

### Median 5' UTRs

```{r}
tmp1 <- dplyr::select(median_adjalley, AssignedFeat, start)
tmp2 <- dplyr::select(fiveutrs, Parent, start, end, width, strand)
compare_adjalley <- dplyr::inner_join(tmp1,tmp2,by=c("AssignedFeat"="Parent")) %>%
  dplyr::mutate(adjalley_width=ifelse(strand=="+",end-start.x,start.x-end))

g <- ggplot(compare_adjalley,aes(adjalley_width+1,width+1)) +
  geom_point(alpha=0.3,size=3.0) +
  geom_abline(color="red") +
  xlab("") + 
  ylab("") +
  scale_x_continuous(trans="log10",breaks=c(1,10,100,1000,10000),
                     labels=scales::trans_format('log10',math_format(10^.x)),limits=c(1,10000)) +
  scale_y_continuous(trans="log10",breaks=c(1,10,100,1000,10000),
                     labels=scales::trans_format('log10',math_format(10^.x)),limits=c(1,10000)) +
  scatter_theme

cowplot::save_plot(filename="../output/compare/adjalley_median_scatter.png",plot=g)
cowplot::save_plot(filename="../output/compare/adjalley_median_scatter.svg",plot=g)
print(g)

g <- ggplot(compare_adjalley,aes(adjalley_width-width)) + 
  geom_histogram(color="grey70",fill="black",bins=25) +
  geom_vline(xintercept=0,color="red",linetype=2,size=1.0) +
  xlab("") +
  ylab("") +
  hist_theme

cowplot::save_plot(filename=paste0("../output/compare/adjalley_median_histogram.png"),plot=g)
cowplot::save_plot(filename=paste0("../output/compare/adjalley_median_histogram.svg"),plot=g)
print(g)

cor.test(compare_adjalley$adjalley_width,compare_adjalley$width)
compare_adjalley %>% dplyr::mutate(diff=adjalley_width-width) %>% summary() %>% print()
```

### Varying thresholds

```{r}
for (i in 1:8) {
  adjalley_filtered <- adjalley %>%
    dplyr::mutate(position=(start+end)/2) %>%
    dplyr::filter(FilterSize>i)
  tmp1 <- dplyr::select(adjalley_filtered, AssignedFeat, position)
  tmp2 <- dplyr::select(utrs, Parent, start)
  compare_adjalley <- dplyr::inner_join(tmp1,tmp2,by=c("AssignedFeat"="Parent"))
  compare_adjalley %>% dplyr::mutate(diff=position-start) %>% summary() %>% print()
  g <- ggplot(compare_adjalley,aes(position-start)) + 
    geom_histogram(color="grey70") +
    xlab("") + 
    ylab("")
  cowplot::save_plot(filename=paste0("../output/compare/adjalley_histogram_",i,".png"),plot=g)
  cowplot::save_plot(filename=paste0("../output/compare/adjalley_histogram_",i,".svg"),plot=g)
  print(g)
}
rm(tmp1,tmp2)
```

