---
title: "Extensive transcription"
author: "Philipp Ross"
date: 2017-03-02
output: workflowr::wflow_html
---

```{r read-chunk, include=FALSE, cache=FALSE}
knitr::read_chunk("chunks.R")
source("../code/utils.R")
load_R_essentials()
```


**Last updated:** `r Sys.Date()`

**Code version:** `r workflowr::extract_commit(".", 1)$sha1`

## Overview

The falciparum genomes sequences show extensive amounts of transcription. A very high percentage of the genome is transcribed, measured here by the number of base pairs covered by at least 5 reads.

## Workflow

1. Calculate genome wide coverage for each individual nucleotide
2. Plot the cumulative distribution for each strain

### Generate genome coverage data

```
for strain in 3d7 hb3 it;
do
  bedtools genomecov -ibam "data/bam/mapped_to_3d7/${strain}.3d7_v3_chr.idc.bam" \
  -g "data/genomes/pf3d7/PlasmoDB-28_Pfalciparum3D7_Genome.lengths" > "output/extensive_transcription/${strain}_genomecov.tsv"
done
```

### Plot data

```{r read-data}
read_genome_cov <- function(file,input_strain) {
  df <- read_tsv(file, col_names=c("seqid","cov","nbases","length","fraction")) %>%
    filter(seqid == "genome") %>% 
    arrange(desc(cov)) %>% 
    mutate(cum = cumsum(fraction))
  df$strain <- input_strain
  return(df)
}

cov3d7 <- read_genome_cov("../output/extensive_transcription/3d7_idc_genomecov.tsv", "3D7")
covhb3 <- read_genome_cov("../output/extensive_transcription/hb3_idc_genomecov.tsv", "HB3")
covit <- read_genome_cov("../output/extensive_transcription/it_idc_genomecov.tsv", "IT4")
cov <- rbind(cov3d7,covhb3,covit)
```

```{r cov-dist}
g <- cov %>% 
  ggplot(aes(x=cov,y=cum,color=strain)) + 
  geom_line(size=1) + 
  scale_x_continuous(trans="log10", breaks=c(1,10,100,1000,10000,100000),labels=trans_format('log10',math_format(10^.x))) +
  outline_colors +
  labs(x="Coverage",y="Cumulative Genome Fraction") +
  geom_vline(xintercept=5,linetype=2)
ggsave(filename="../output/extensive_transcription/extensive_transcription.pdf",plot=g)
```

```{r}
print(g)
```

## Session Information

