---
title: "Differential phase detection"
author: "Philipp Ross"
date: 2017-04-17
output: html_document
---

```{r read-chunk, include=FALSE, cache=FALSE}
knitr::read_chunk("chunks.R")
```

```{r knitr-opts-chunk, include=FALSE}
```

```{r load-libraries, include=FALSE}
```

**Last updated:** `r Sys.Date()`

**Code version:** `r workflowr::extract_commit(".", 1)$sha1`

## Calculating the phases for each gene

First let's import and process the data we need to use for this:

```{r}
# import expression data
x3d7exp <- readRDS("../output/neighboring_genes/gene_reduced_3d7_abund.rds")
xhb3exp <- readRDS("../output/neighboring_genes/gene_reduced_hb3_abund.rds")
xitexp  <- readRDS("../output/neighboring_genes/gene_reduced_it_abund.rds")
exp <- dplyr::bind_rows(x3d7exp,xhb3exp,xitexp)

rm(x3d7exp,xhb3exp,xitexp)

# import core genes for comparison
core_genes <- readr::read_tsv("../data/gene_lists/core_pf3d7_genes.txt",col_names=F)$X1

# only consider genes with a TPM of 5 within at least one sample
detected  <- exp %>%
  dplyr::filter(gene_id %in% core_genes) %>%
  dplyr::group_by(gene_id, strain) %>%
  dplyr::summarise(n = sum(TPM >= 5)) %>%
  dplyr::mutate(on = ifelse(n > 0, 1, 0)) %>%
  dplyr::ungroup() %>%
  dplyr::select(-n)

on3d7 <- detected %>% dplyr::filter(strain == "3d7" & on == 1) %$% gene_id
onhb3 <- detected %>% dplyr::filter(strain == "hb3" & on == 1) %$% gene_id
onit  <- detected %>% dplyr::filter(strain == "it" & on == 1) %$% gene_id

on3d7hb3 <- intersect(on3d7,onhb3)
on3d7it  <- intersect(on3d7,onit)
onhb3it  <- intersect(onhb3,onit)
# create a list of all genes on between strains
onall <- Reduce(intersect, list(on3d7, onhb3, onit))
```

We also only want to consider genes with a decent amplitude to remove genes that are constitutively on:

```{r}
amps <- abund %>%
  dplyr::filter(gene_id %in% onall) %>%
  dplyr::group_by(gene_id, strain) %>%
  dplyr::summarise(min=min(TPM),max=max(TPM)) %>%
  dplyr::mutate(amp=max/(min+1)) %>%
  dplyr::ungroup()

amps <- amps %>% dplyr::inner_join(as.data.frame(org.Pf.plasmoGENENAME), by = "gene_id")

amps %>%
  dplyr::mutate(strain = toupper(strain)) %>%
  ggplot(aes(x=amp,group=strain,color=strain)) +
  geom_line(stat="density",size=1) +
  fill_colors +
  outline_colors +
  xlab("Amplitude") +
  ylab("Density") + 
  scale_x_continuous(trans="log10", breaks=c(1,10,100,1000,10000),labels=trans_format('log10',math_format(10^.x))) +
  theme(legend.title=element_blank(),axis.line.x=element_line(size=0),axis.line.y=element_line(size=0)) +
  panel_border(colour="black")

high_amp <- amps %>% dplyr::filter(log2(amp) > 0.5) %$% unique(gene_id)
```

In Broadbent et al. they generate a circular plot to represent the different timepoints as x and y coordinates following dimensional reduction using multidimensional scaling. Prior to this we need to correct for impure Plasmodium cell populations as demonstrated in the following plot:

```{r}
wide_uncorrected <- exp %>% 
  dplyr::filter(gene_id %in% onall & gene_id %in% high_amp) %>%
  dplyr::group_by(gene_id,strain) %>% 
  dplyr::summarise(m=mean(TPM)) %>% 
  dplyr::inner_join(exp) %>% 
  dplyr::mutate(norm_tpm=(((TPM/m)-mean(TPM/m))/sd(TPM/m))) %>% 
  dplyr::select(gene_id,tp,norm_tpm,strain) %>% 
  dplyr::ungroup() %>%
  tidyr::unite(tp_strain, tp, strain, sep = "-") %>%
  dplyr::select(tp_strain, norm_tpm, gene_id) %>%
  tidyr::spread(tp_strain, norm_tpm) %>%
  dplyr::select(-gene_id)

d <- as.dist((1 - cor(wide_uncorrected))/2)

fit <- MASS::isoMDS(d, k = 2)

timepoint_phase <- as.data.frame(fit$points)
timepoint_phase$rows <- rownames(fit$points)
timepoint_phase <- timepoint_phase %>%
  tidyr::separate(col=rows, into=c("tp","strain"), sep="-") %>%
  dplyr::rename(x=V1,y=V2)

write_tsv(x = timepoint_phase, path = "../output/differential_phase/uncorected_timepoint_phase.tsv" )

timepoint_phase %>% 
  ggplot(aes(x=x,y=y,label=tp,color=toupper(strain))) + 
  geom_point() + 
  geom_text(check_overlap=F,nudge_x=0.05) +
  ggtitle("Non-Corrected MDS Plot") +
  panel_border(colour="black",size=1) +
  theme(panel.grid = element_line(size=1)) +
  outline_colors
```

To correct for the phases we can do the following:

```{r}
array_3d7 <- read.delim ("../data/array_correlations/selected_3d7array_foldchanges_v3_geneids.txt")

df1 <- exp %>% 
  dplyr::filter(gene_id %in% onall & gene_id %in% high_amp) %>%
  dplyr::group_by(gene_id,strain) %>% 
  dplyr::summarise(m=mean(TPM)) %>% 
  dplyr::inner_join(exp) %>% 
  dplyr::mutate(norm_tpm=(((TPM/m)-mean(TPM/m))/sd(TPM/m))) %>% 
  dplyr::select(gene_id,tp,norm_tpm,strain) %>% 
  dplyr::ungroup() %>%
  dplyr::rowwise() %>%
  dplyr::mutate(hpi=ifelse(tp==1,1,if(tp==2){8}else if(tp==3){16}else if(tp==4){24}else if(tp==5){32}else if(tp==6){40}else{48})) %>%
  dplyr::ungroup() %>%
  dplyr::do(fit = predict(splines::periodicSpline(norm_tpm ~ hpi, ., period = 48), seq(1,53,1))) %>%
  broom::tidy(fit) %>%
  dplyr::ungroup() %>%
  dplyr::rename(tp=x,exp=y)

dups <- as.character(array_3d7$GeneID[duplicated(array_3d7 %$% GeneID)])

df2 <- array_3d7 %>%
  filter(GeneID %nin% dups) %>%
  gather(tp,exp,-GeneID) %>%
  dplyr::rename(gene_id = GeneID)

df1_3d7 <- df1 %>% filter(strain == "3d7") %>% dplyr::select(-strain)
df1_hb3 <- df1 %>% filter(strain == "hb3") %>% dplyr::select(-strain)
df1_it <- df1 %>% filter(strain == "it") %>% dplyr::select(-strain)

df2$gene_id <- as.character(df2$gene_id)
df2 <- df2 %>% mutate(tp = ifelse(as.numeric(str_replace_all(tp, "Timepoint.","")) > 9, paste0("array_",str_replace_all(tp, "Timepoint.","")), paste0("array_0",str_replace_all(tp, "Timepoint.",""))))
df1_3d7$tp <- ifelse(df1_3d7$tp > 9, paste(paste0("rnaseq_",df1_3d7$tp)), paste(paste0("rnaseq_0",df1_3d7$tp)))
df1_hb3$tp <- ifelse(df1_hb3$tp > 9, paste(paste0("rnaseq_",df1_hb3$tp)), paste(paste0("rnaseq_0",df1_hb3$tp)))
df1_it$tp <- ifelse(df1_it$tp > 9, paste(paste0("rnaseq_",df1_it$tp)), paste(paste0("rnaseq_0",df1_it$tp)))

wide2 <- df2 %>% spread(tp,exp)
```

```{r}
wide1 <- df1_3d7 %>% spread(tp,exp)

joined <- inner_join(wide1,wide2, by = "gene_id")

c3d7 <- max.col(t(cor(joined[,2:54], joined[,55:107], use = "na.or.complete")))

wide1 <- wide1[c(1,c3d7[c(1,8,16,24,32,40,48)])]
colnames(wide1) <- c("gene_id",1,8,16,24,32,40,48)

exp_3d7 <- wide1 %>%
  gather(tp,exp,-gene_id) %>%
  group_by(gene_id) %>%
  do(fit = predict(splines::periodicSpline(exp ~ tp, ., period = 48), seq(1,48,1))) %>%
  tidy(fit) %>%
  ungroup()

exp_3d7 %<>% mutate(strain = "3d7") %>% dplyr::rename(tp = x, exp = y)


wide1 <- df1_hb3 %>% spread(tp,exp)

joined <- inner_join(wide1,wide2, by = "gene_id")

chb3 <- max.col(t(cor(joined[,2:54], joined[,55:107], use = "na.or.complete")))

wide1 <- wide1[c(1,chb3[c(1,8,16,24,32,40,48)])]
colnames(wide1) <- c("gene_id",1,8,16,24,32,40,48)

exp_hb3 <- wide1 %>%
  gather(tp, exp,-gene_id) %>%
  group_by(gene_id) %>%
  do(fit = predict( splines::periodicSpline(exp ~ tp, ., period = 48), seq(1,48,1))) %>%
  tidy(fit) %>%
  ungroup()

exp_hb3 %<>% mutate(strain = "hb3") %>% dplyr::rename(tp = x, exp = y)


wide1 <- df1_it %>% spread(tp,exp)

joined <- inner_join(wide1,wide2, by = "gene_id")

cit <- max.col(t(cor(joined[,2:54], joined[,55:107], use = "na.or.complete")))

wide1 <- wide1[c(1,cit[c(1,8,16,24,32,40,48)])]
colnames(wide1) <- c("gene_id",1,8,16,24,32,40,48)

exp_it <- wide1 %>%
  gather(tp, exp,-gene_id) %>%
  group_by(gene_id) %>%
  do(fit = predict( splines::periodicSpline(exp ~ tp, ., period = 48), seq(1,48,1))) %>%
  tidy(fit) %>%
  ungroup()

exp_it %<>% mutate(strain = "it") %>% dplyr::rename(tp = x, exp = y)

final <- bind_rows(exp_3d7,exp_hb3,exp_it) %>% filter(tp %in% c(1,8,16,24,32,40,48))

wide <- final %>% 
  unite(tp_strain, tp, strain) %>% 
  dplyr::select(gene_id, tp_strain, exp) %>% 
  spread(tp_strain, exp) %>% 
  dplyr::select(-gene_id)

d <- as.dist((1 - cor(wide))/2)

fit <- isoMDS(d+0.0000001, k = 2)

dfit <- as.data.frame(fit$points)
dfit$tp_strain <- rownames(dfit)
rownames(dfit) <- NULL

dfit %>% 
  separate(tp_strain, c("tp" ,"strain"), "_") %>% 
  ggplot(aes(x=V1,y=V2,label=tp,color=toupper(strain))) + 
  geom_point() + 
  geom_text(check_overlap=F,nudge_x=0.05) +
  ggtitle("Locally Shifted MDS Plot") +
  panel_border(colour="black",size=1) +
  theme(panel.grid = element_line(size=1)) +
  outline_colors

```

```{r,eval=F}
wide <- exp %>% 
  dplyr::filter(gene_id %in% onall & gene_id %in% high_amp) %>%
  dplyr::group_by(gene_id,strain) %>% 
  dplyr::summarise(m=mean(TPM)) %>% 
  dplyr::inner_join(exp) %>% 
  dplyr::mutate(norm_tpm=(((TPM/m)-mean(TPM/m))/sd(TPM/m))) %>% 
  dplyr::select(gene_id,tp,norm_tpm,strain) %>% 
  dplyr::ungroup() %>%
  tidyr::unite(gene_strain, gene_id, strain, sep = "-") %>%
  dplyr::select(gene_strain, norm_tpm, tp) %>%
  tidyr::spread(gene_strain, norm_tpm)

d <- as.dist((1 - cor(wide))/2)

fit <- MASS::isoMDS(d, k = 2)

gene_phase <- as.data.frame(fit$points)
gene_phase$rows <- rownames(fit$points)
gene_phase <- gene_phase %>%
  tidyr::separate(col=rows, into=c("gene_id","strain"), sep="-") %>%
  dplyr::rename(x=V1,y=V2)

gene_phase$phase <- apply(gene_phase[,1:2], 1, function(x) {atan2(x[1], x[2])})
rownames(gene_phase) <- NULL

gene_phase <- gene_phase %>% 
  dplyr::select(gene_id,strain,x,y,phase)

readr::write_tsv(x = gene_phase, path = "../output/differential_phase/gene_phase.tsv" )
```

## Phase comparison histograms

```{r}
# Prepare phase shifts between strains
s <- gene_phase %>%
  dplyr::select(gene_id, strain, phase) %>%
  tidyr::spread(strain, phase) %>%
  dplyr::rowwise() %>%
  dplyr::mutate(`3D7 - HB3` = min(c(abs(`3d7`-hb3),2*pi-abs(`3d7`-hb3))),
         `3D7 - IT` = min(c(abs(`3d7`-it),2*pi-abs(`3d7`-it))),
         `HB3 - IT` = min(c(abs(hb3-it),2*pi-abs(hb3-it))))

# calculate median that 3D7 is off by
m <- s %>%
  rowwise() %>%
  mutate(offset = min(c(abs(`3d7`-mean(c(hb3,it))),2*pi-abs(`3d7`-mean(c(hb3,it)))))) %$%
  median(offset)

# Recalculate phase shifts
s <- s %>%
  mutate(`3d7` = ifelse(`3d7` > 0, `3d7` - m, `3d7` + m)) %>%
  mutate(`3D7 - HB3` = min(c(abs(`3d7`-hb3),2*pi-abs(`3d7`-hb3))),
         `3D7 - IT` = min(c(abs(`3d7`-it),2*pi-abs(`3d7`-it))),
         `HB3 - IT` = min(c(abs(hb3-it),2*pi-abs(hb3-it))))

# plot histograms
s %>%
  ungroup() %>%
  filter(gene_id %in% onall) %>%
  dplyr::select(-`3d7`,-hb3,-it) %>%
  gather(strains,diff,-gene_id) %>%
  filter(strains == "3D7 - HB3") %>%
  ggplot(aes(x=diff)) +
  geom_histogram(color="grey70") +
  xlab("Phase Difference") +
  ylab("Frequency") +
  panel_border(colour="black",size=1) +
  scale_x_continuous(limits=c(0,3),breaks=c(0,pi/2,pi),labels=c("0","pi/2","pi")) +
  theme(axis.line.x=element_line(size=0),axis.line.y=element_line(size=0)) +
  ggtitle("3D7 vs. HB3")

s %>%
  ungroup() %>%
  filter(gene_id %in% onall) %>%
  dplyr::select(-`3d7`,-hb3,-it) %>%
  gather(strains,diff,-gene_id) %>%
  filter(strains == "3D7 - IT") %>%
  ggplot(aes(x=diff)) +
  geom_histogram(color="grey70") +
  xlab("Phase Difference") +
  ylab("Frequency") +
    panel_border(colour="black",size=1) +
  scale_x_continuous(limits=c(0,3),breaks=c(0,pi/2,pi),labels=c("0","pi/2","pi")) +
  theme(axis.line.x=element_line(size=0),axis.line.y=element_line(size=0)) +
  ggtitle("3D7 vs. IT")

s %>%
  ungroup() %>%
  filter(gene_id %in% onall) %>%
  dplyr::select(-`3d7`,-hb3,-it) %>%
  gather(strains,diff,-gene_id) %>%
  filter(strains == "HB3 - IT") %>%
  ggplot(aes(x=diff)) +
  geom_histogram(color="grey70") +
  xlab("Phase Difference") +
  ylab("Frequency") +
    panel_border(colour="black",size=1) +
  scale_x_continuous(limits=c(0,3),breaks=c(0,pi/2,pi),labels=c("0","pi/2","pi")) +
  theme(axis.line.x=element_line(size=0),axis.line.y=element_line(size=0)) +
  ggtitle("HB3 vs. IT")
  

# plot scatterplots
s %>%
  filter(gene_id %in% onall) %>%
  ggplot(aes(y=`3d7`,x=hb3)) +
  geom_point(alpha=0.50,color="grey50") +
  scale_x_continuous(breaks = c()) +
  scale_y_continuous(breaks = c()) +
  geom_hline(yintercept = 0) +
  geom_vline(xintercept = 0) +
  theme(axis.line = element_blank())

s %>%
  filter(gene_id %in% onall) %>%
  ggplot(aes(y=`3d7`,x=it)) +
  geom_point(alpha=0.50,color="grey50") +
  scale_x_continuous(breaks = c()) +
  scale_y_continuous(breaks = c()) +
  geom_hline(yintercept = 0) +
  geom_vline(xintercept = 0) +
  theme(axis.line = element_blank())

s %>%
  filter(gene_id %in% onall) %>%
  ggplot(aes(y=hb3,x=it)) +
  geom_point(alpha=0.50,color="grey50") +
  scale_x_continuous(breaks = c()) +
  scale_y_continuous(breaks = c()) +
  geom_hline(yintercept = 0) +
  geom_vline(xintercept = 0) +
  theme(axis.line = element_blank())
```

## Session Information

```{r session-info}
```
