---
title: "Differential transcript detection"
author: "Philipp Ross"
date: 2017-04-17
output: html_document
---

```{r read-chunk, include=FALSE, cache=FALSE}
knitr::read_chunk("chunks.R")
```

```{r knitr-opts-chunk, include=FALSE}
```

```{r load-libraries, include=FALSE}
```

**Last updated:** `r Sys.Date()`

**Code version:** `r workflowr::extract_commit(".", 1)$sha1`

## What transcripts can we detect in one strain, but not the others?

First we need to detect transcript differences. To do this we run the `detect_transcripts.R` script found within `code/differential_detection` directory.

## Are these genes enriched for any GO terms?

Next we run the topGO script:

```{r,engine='bash'}
bash code/differential_detection/run_topgo.sh data/gene_ontology/Pf3D7_go_sept2014.txt data/gene_ontology/Pf3D7_desc_sept2014.txt
```

## Are undetected genes sometimes due to known polymorphic regions?



## Results

Some genes may barely get over the threshold of 5 RPKMs, which was chosen arbitrarily. We are not interested in those results. We want to see what transcripts are clearly abundant in one strain, while clearly absent or at the very least severely down-regulated in another strain. The question then becomes whether this is due to poor coverage across the gene (highly polymorphic) or actual low transcript copy number. MSP1 is an example of a highly polymorphic gene in HB3 versus 3D7. It is not covered very well across the gene and thus has been marked as "undetected." MSP2 is an example of a gene that is detected, however, also highly polymorphic.

Looking through this CGH data will help as well: http://www.ncbi.nlm.nih.gov/pmc/articles/PMC2901668/

####Up in 3D7

PF3D7_0202000 2145.335 versus 2.995
knob-associated histidine-rich protein (KAHRP)

PF3D7_1372200 836.830 versus 0.599
histidine-rich protein III (HRPIII)

PF3D7_0201700 110.712 versus 0.052
DnaJ protein, putative

PF3D7_1429200 24.382 versus 0.99
transcription factor with AP2 domain(s)

PF3D7_0518700 30.287 versus 1.4
mRNA-binding protein PUF1 (PUF1)

PF3D7_1126800 37.441 versus 2.2023
RNA-binding protein, putative

PF3D7_1350900 43.551 versus 1.301
transcription factor with AP2 domain(s) (ApiAP2)

PF3D7_1222600 50.469 versus 0.951
transcription factor with AP2 domain(s) (AP2-G)

PF3D7_1241400 19.426 versus 1.633
RNA-binding protein, putative

PF3D7_0209000 34.424 versus 0.084
6-cysteine protein (P230)

PF3D7_1346700 46.211 versus 2.543
6-cysteine protein (P48/45)

PF3D7_1346800 13.754 versus 1.110
6-cysteine protein (P47)

PF3D7_1038400 14.784 versus 0.030
gametocyte-specific protein (Pf11-1)

PF3D7_0625600 31.212 versus 2.054
poly(A) polymerase PAP, putative

PF3D7_1031000 27.841 versus 0.468
25 kDa ookinete surface antigen precursor (Pfs25)

####Up in HB3

10 out of top 30 are pseudogenes

Not many differences beyond highly variant genes (variant due to clonal variation or polymorphic regions? Both?)

stevor, exported protein family 4, pfmc-2tm mauler’s cleft two transmembrane protein, mspr4, g-beta repeat protein

###3D7 and IT

####Up in 3D7

See a lot of the same genes here as I did in 3D7 versus HB3.

PF3D7_1222600
transcription factor with AP2 domain(s) (AP2-G)

PF3D7_1346700
6-cysteine protein (P48/45)

PF3D7_0209000
6-cysteine protein (P230)

PF3D7_1477800
acyl-CoA binding protein (ACBP)

PF3D7_0414500
RNA-binding protein, putative

PF3D7_1031000
25 kDa ookinete surface antigen precursor (Pfs25)

PF3D7_1429200
transcription factor with AP2 domain(s)

PF3D7_0311700
plasmepsin VI

PF3D7_0109300
fatty acid elongation protein, GNS1/SUR4 family, putative

PF3D7_1038400
gametocyte-specific protein (Pf11-1)

####Up in IT

PF3D7_0114300
exported protein family 4, pseudogene (EPF4)

PF3D7_0936700
lysophospholipase, putative

###HB3 and IT

Not much difference here.

####Up in HB3

PF3D7_1477800
acyl-CoA binding protein (ACBP)

####Up in IT

PF3D7_0202000
knob-associated histidine-rich protein (KAHRP)

PF3D7_1372200
histidine-rich protein III (HRPIII)

PF3D7_0201800
knob associated heat shock protein 40 (KAHsp40)

PF3D7_0201700
DnaJ protein, putative

PF3D7_0625600
poly(A) polymerase PAP, putative


## Session Information

```{r session-info}
```
