---
title: "Transcription Factor Binding Site Analysis"
author: "Philipp Ross"
date: 2017-07-16
output: html_document
---

```{r read-chunk, include=FALSE, cache=FALSE}
knitr::read_chunk("chunks.R")
```

```{r knitr-opts-chunk, include=FALSE}
```

```{r load-libraries, include=FALSE}
```

**Last updated:** `r Sys.Date()`

**Code version:** `r workflowr::extract_commit(".", 1)$sha1`

## Refined motif search

Now that we have predicted TSSs, we can refine our motif binding site search space and look at smaller regions of sequence just upstream of our predicted TSSs. We will use our most commonly used TSSs for this.

First we need to create promoter regions to analyze:

```{r}
promoter_region <- 1000

promoters_3d7 <- rtracklayer::import.gff3("../output/final_utrs/longest_utrs_3d7_plasmodb_compatible.gff") %>% 
  tibble::as_tibble() %>% 
  dplyr::mutate(Parent=stringr::str_replace(stringr::str_replace(stringr::str_replace(unlist(Parent), "rna_", ""), "-1", ""),"[.][0-9]","")) %>% 
  dplyr::distinct() %>%
  dplyr::mutate(end=ifelse(strand=="+",start,end+promoter_region),start=ifelse(strand=="+",start-promoter_region,end)) %>%
  dplyr::filter(type=="5UTR") %>%
  GenomicRanges::GRanges()

promoters_hb3 <- rtracklayer::import.gff3("../output/final_utrs/longest_utrs_hb3_plasmodb_compatible.gff") %>% 
  tibble::as_tibble() %>% 
  dplyr::mutate(Parent=stringr::str_replace(stringr::str_replace(stringr::str_replace(unlist(Parent), "rna_", ""), "-1", ""),"[.][0-9]","")) %>% 
  dplyr::distinct() %>%
  dplyr::mutate(end=ifelse(strand=="+",start,end+promoter_region),start=ifelse(strand=="+",start-promoter_region,end)) %>%
  dplyr::filter(type=="5UTR") %>%
  GenomicRanges::GRanges()

promoters_it <- rtracklayer::import.gff3("../output/final_utrs/longest_utrs_it_plasmodb_compatible.gff") %>% 
  tibble::as_tibble() %>% 
  dplyr::mutate(Parent=stringr::str_replace(stringr::str_replace(stringr::str_replace(unlist(Parent), "rna_", ""), "-1", ""),"[.][0-9]","")) %>% 
  dplyr::distinct() %>%
  dplyr::mutate(end=ifelse(strand=="+",start,end+promoter_region),start=ifelse(strand=="+",start-promoter_region,end)) %>%
  dplyr::filter(type=="5UTR") %>%
  GenomicRanges::GRanges()

rtracklayer::export.gff3(object=promoters_3d7,con="../output/tfbs_analysis/promoters_3d7.gff")
rtracklayer::export.gff3(object=promoters_hb3,con="../output/tfbs_analysis/promoters_hb3.gff")
rtracklayer::export.gff3(object=promoters_it,con="../output/tfbs_analysis/promoters_it.gff")

seqs3d7 <- BSgenome::getSeq(BSgenome.Pfalciparum.PlasmoDB.v24, promoters_3d7)
names(seqs3d7) <- stringr::str_replace(stringr::str_replace(unlist(promoters_3d7$Parent), "rna_", ""), "-1", "")
Biostrings::writeXStringSet(seqs3d7,"../output/tfbs_analysis/promoters_3d7.fasta")
seqshb3 <- BSgenome::getSeq(BSgenome.Pfalciparum.PlasmoDB.v24, promoters_hb3)
names(seqshb3) <- stringr::str_replace(stringr::str_replace(unlist(promoters_hb3$Parent), "rna_", ""), "-1", "")
Biostrings::writeXStringSet(seqshb3,"../output/tfbs_analysis/promoters_hb3.fasta")
seqsit <- BSgenome::getSeq(BSgenome.Pfalciparum.PlasmoDB.v24, promoters_it)
names(seqsit) <- stringr::str_replace(stringr::str_replace(unlist(promoters_it$Parent), "rna_", ""), "-1", "")
Biostrings::writeXStringSet(seqsit,"../output/tfbs_analysis/promoters_it.fasta")
```

We should also create files that parse the genomic coordinates so as to be able to load them into a genome browser:

```{r,engine='bash',eval=F}
for strain in 3d7 hb3 it; do bedtools getfasta -fi ../data/genomes/pf3d7/PlasmoDB-24_Pfalciparum3D7_Genome.fasta -bed ../output/tfbs_analysis/promoters_it.gff -fo ../output/tfbs_analysis/promoters_it_genome_coords.fasta -s; done
```

Now we need to run fimo on these regions and search for our motifs:

```{r, engine='bash',eval=F}
# generate background files
for strain in 3d7 hb3 it; do fasta-get-markov -m 3 ../output/tfbs_analysis/promoters_${strain}.fasta ../output/tfbs_analysis/promoters_${strain}.background; done
# run fimo for promoters with gene names
for strain in 3d7 hb3 it; do fimo --bgfile ../output/tfbs_analysis/promoters_${strain}.background --no-qvalue -oc ../output/tfbs_analysis/promoters_${strain} --thresh 1e-2 ../data/motif_cores/ap2_pbm_cores.meme ../output/tfbs_analysis/promoters_${strain}.fasta; done
# and for promoters with genome coordinates
for strain in 3d7 hb3 it; do fimo --bgfile ../output/tfbs_analysis/promoters_${strain}.background --no-qvalue --parse-genomic-coord -oc ../output/tfbs_analysis/promoters_${strain}_genome_coords --thresh 1e-2 ../data/motif_cores/ap2_pbm_cores.meme ../output/tfbs_analysis/promoters_${strain}_genome_coords.fasta; done
```

Now we can import that data and analyze it for positional information:

```{r}
motifs_3d7 <- rtracklayer::import.gff3("../output/tfbs_analysis/promoters_3d7/fimo.gff") %>% tibble::as_tibble()

for (name in unique(motifs_3d7$Name)) {
  n <- motifs_3d7 %>% 
    dplyr::filter(Name==name)
  #print(ks.test(n$start,runif(10000),alternative="less"))
  g <- n %>% ggplot(aes(x=start)) + geom_histogram(color="grey70") + xlab("Start Site") + ylab("Frequency")
  ggsave(plot=g,filename=paste0("../output/tfbs_analysis/promoters_3d7/",name,".pdf"))
}

motifs_hb3 <- rtracklayer::import.gff3("../output/tfbs_analysis/promoters_hb3/fimo.gff") %>% tibble::as_tibble()

for (name in unique(motifs_hb3$Name)) {
  n <- motifs_hb3 %>% 
    dplyr::filter(Name==name)
  #print(ks.test(n$start,runif(10000),alternative="less"))
  g <- n %>% ggplot(aes(x=start)) + geom_histogram(color="grey70")
  ggsave(plot=g,filename=paste0("../output/tfbs_analysis/promoters_hb3/",name,".pdf"))
}

motifs_it <- rtracklayer::import.gff3("../output/tfbs_analysis/promoters_it/fimo.gff") %>% tibble::as_tibble()

for (name in unique(motifs_it$Name)) {
  n <- motifs_it %>% 
    dplyr::filter(Name==name)
  #print(ks.test(n$start,runif(10000),alternative="less"))
  g <- n %>% ggplot(aes(x=start)) + geom_histogram(color="grey70")
  ggsave(plot=g,filename=paste0("../output/tfbs_analysis/promoters_it/",name,".pdf"))
}
```

## ApiAP2 Activity Predictions

Here we will redo the analysis performed in Campbell et al. using motif hits within newly defined promoter regions. To perform this analysis, two R scripts need to be sourced from within the analysis working directory.

First `generate_inputs.R`, then `estimate_apiap2_activity.R`.

## Bidirectional promoters

It would be interesting to see whether there is an enrichment for a particular motif found within bidirectional promoters. First let's extract bidirectional promoter sequences:

```{r}
for (strain in c("3d7","hb3","it")) {

divergent <- readr::read_tsv(paste0("../output/neighboring_genes/",strain,"_divergent.tsv")) %>%
  dplyr::filter(dist < 1000 & dist > 0 & cor >= 0.5)

transcripts <- tibble::as_tibble(rtracklayer::import.gff3(paste0("../output/neighboring_genes/full_transcripts_",strain,".gff")))
biprom <- tibble::tibble(seqnames=character(),start=integer(),end=integer(),strand=character(),source=character(),type=character(),ID=character())

for (i in 1:nrow(divergent)) {
  start_gene <- dplyr::filter(transcripts, ID == divergent$left_gene[i])
  end_gene <- dplyr::filter(transcripts, ID == divergent$right_gene[i])
  new <- tibble::tibble(seqnames=start_gene$seqnames,
                        start=start_gene$end,
                        end=end_gene$start,
                        strand="+",
                        source="PlasmoDB",
                        type="BiPromoter",
                        ID=paste0(start_gene$ID,"-",end_gene$ID))
  biprom <- dplyr::bind_rows(biprom, new)
}
biprom <- GenomicRanges::GRanges(biprom)
rtracklayer::export.gff3(object=biprom,con=paste0("../output/tfbs_analysis/bidirectional_",strain,".gff"))
seqs <- BSgenome::getSeq(BSgenome.Pfalciparum.PlasmoDB.v24, biprom)
names(seqs) <- biprom$ID
Biostrings::writeXStringSet(seqs,paste0("../output/tfbs_analysis/bidirectional_",strain,".fasta"))
}
```

Now create background files and run fimo:

```{r, engine='bash',eval=F}
for strain in 3d7 hb3 it; do fasta-get-markov -m 3 ../output/tfbs_analysis/bidirectional_${strain}.fasta ../output/tfbs_analysis/bidirectional_${strain}.background; done
for strain in 3d7 hb3 it; do fimo --bgfile ../output/tfbs_analysis/bidirectional_${strain}.background --no-qvalue -oc ../output/tfbs_analysis/bidirectional_${strain} --thresh 1e-2 ../data/motif_cores/ap2_pbm_cores.meme ../output/tfbs_analysis/bidirectional_${strain}.fasta; done
```

And now we can read those files in and check the motif occurences:

```{r}
bimotifs_3d7 <- rtracklayer::import.gff3("../output/tfbs_analysis/bidirectional_3d7/fimo.gff") %>% tibble::as_tibble()

df <- tibble::as_tibble(biprom_3d7)
newdf <- tibble::tibble(name=character(),norm_start=numeric())

for (m in unique(bimotifs_3d7$seqnames)) {
  w <- dplyr::filter(df, ID == m)$width
  newdf <- dplyr::bind_rows(newdf, dplyr::filter(bimotifs_3d7, seqnames == m) %>% dplyr::transmute(name=seqnames, norm_start = start / w, motif=Name))
}

for (n in unique(newdf$motif)) {
  b <- newdf %>% dplyr::filter(motif==n)
  g <- b %>% ggplot(aes(x=norm_start)) + geom_line(stat="density")
  ggsave(plot=g,filename=paste0("../output/tfbs_analysis/bidirectional_3d7/",n,".pdf"))
}
```

## Dynamic motif usage

Additionally, we can look at TSSs that we can confidentally say is shifting and analyze the motifs within these regions to find a nice example to display. We did this for KARHP:




## Session Information

```{r session-info}
```
